Higher-Order unification for free!

### 2024-04-12 paper/main.tex
typo
section 2 done (can be spellchecked)

\noindent
The main point in showing these equality tests is to remark how
weaker \Ee is, and to identify the four rules that need special
treatment in the implementation of \Uo.
For brevity we omit the code of \elpiIn{beta}:
it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
weak head normal form of <<\elpiIn{app [F|L]}>>.
Note that the symbol \elpiIn{|} separates the head of a list from the tail.

@@

\subsection{The problem: logic-program simulation}
% \todo{dire che P e T sono liste di coppie, dire che P e T vengono anche usati come insiemi}
We represent a logic program \emph{run} in \Fo{} as
a list \emph{steps} of length $\mathcal{N}$.
\foUnifPb is a list of pair of terms, each pair corresponds to a unification
problem. At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$ (the small $l$ and $r$ represent respectively the left and right 
elements extracted from $\foUnifPb_p$).
\footnote{If the same rule is used multiple times in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time.}
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, that is the result of the
logic-program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

\noindent
\todo{say that backtracking is not important} 
In order to simulate a \Fo{} logic program in \Ho{} we compile
each pair of \Fo{} terms in \foUnifPb into a pair of \Ho{} terms $t$.
We write this translation $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produce a variable mapping $m$ and list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.

We simulate each run in \Fo{} with a run in \Ho as follows.
%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\end{array}
$$
$$
\begin{array}{l}
  \runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t,m,l) | s \in \foUnifPb, \C{s}{t}{m}{l} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

\noindent
By analogy with \foUnifPb, we write $\hoUnifPb_{p_l}$ and $\hoUnifPb_{p_r}$
for the two \Ho{} terms being unified at step $p$, and we write $\hoUnifPb_p$
for the pair $(\hoUnifPb_{p_l}, \hoUnifPb_{p_r})$.\\
\hstep{} is made of two sub-steps: a call to the meta language
unification and a check for \progress{} on the set of links, that intuitively
will compensate for the weaker equational theory honoured by \Ue.
\hrun{} compiles all terms in \foUnifPb{}, then executes each step and
finally decompiles the solution.
We claim:

@@

\noindent
That is, the two executions give the same result. Moreover:

@@

In other words if the two terms involved in a step
re-enter \llambda, then \hstep and \fstep are again related, even if
$\foUnifPb \not\subseteq \llambda$ and hence \cref{prop:fidelity} does not apply.

This property has a practical relevance since in many logic programming
implementations, including Elpi, the order in which unification problems
are tackled does matter.
The simplest example is the sequence $F \UnifRel \lambda x.a$ and
$F \appsep a \UnifRel a$: the second problem is not in \llambda and has two
unifiers, namely $\sigma_1 = \{~ F \mapsto \lambda x.x ~\}$ and
$\sigma_2 = \{~ F \mapsto \lambda x.a ~\}$. The first problem picks $\sigma_2$
making the second problem re-enter \llambda.
\todo{say that this is runtime, while the other fidelity is compile time}

\subsection{The solution (in a nutshell)}
\label{sec:nutshell}
A term $s$ is compiled to a term $t$ where every
``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
with an accessory \emph{link} that represents a suspended unification problem
$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, in the sense that
it does not contradict \Eo as it would otherwise do on the
``problematic'' sub-terms.\\
We now define ``problematic'' and ``well behaved'' more formally
and we pick the $\Diamond$ symbol since it
stands for ``possibly'' in modal logic and all problematic terms
are characterized by some ``uncertainty''.

@@

\noindent
\todo{dire perchè W(T), dire che il converse non vale (alcuni check li fa la progress)}
Note that the property holds for any substitution, it could be given by an
oracle and not necessarily a most general one, and for any terms,
in particular terms in $\maybebeta{} ~\cup~ \maybeeta{} ~\cup~ \notllambda{}$.

>>>>

\noindent
The main point in showing these equality tests is to remark how
weaker \Ee is, and to identify the four rules that need special
treatment in the implementation of \Uo.
For brevity we omit the code of \elpiIn{beta}:
it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
weak head normal form of <<\elpiIn{app [F|L]}>>.
Note that the symbol \elpiIn{|} separates the head of a list from the tail.
\todo{explain forall2}

@@

\subsection{The problem: logic-program simulation}
We represent a logic program \emph{run} in \Fo{} as
a sequence of \emph{steps} of length $\mathcal{N}$.
At each step $p$ we unify two terms, $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$, taken from the list of all unification
problems \foUnifPb.
% \footnote{If the same rule is used multiple times in a run we
% consider as many copies as needed of the terms composing the
% rules, with fresh unification variables each time. Also, we do not
% model backtracking directly, but rather the fact that a step in a run may
% fail, and in that case the entire run is said to fail. A regular logic program
% that succeeds is modeled as a many runs, with possibly overlapping starts,
% where at least one run does not end in a failure. }
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, that is the result of the
logic-program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

\noindent
\todo{say that backtracking is not important} 
In order to simulate a \Fo{} logic program in \Ho{} we compile
each \Fo{} term $s$ in \foUnifPb to a \Ho{} term $t$.
We write this translation $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produce a variable mapping $m$ and list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.
We write $\hoUnifPb_p = \{~ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} ~\}$
and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.

We simulate each run in \Fo{} with a run in \Ho as follows:

%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
  \runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t,m,l) | s \in \foUnifPb, \C{s}{t}{m}{l} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

\noindent
By analogy with \foUnifPb, we write $\hoUnifPb_{p_l}$ and $\hoUnifPb_{p_r}$
for the two \Ho{} terms being unified at step $p$, and we write $\hoUnifPb_p$
for the set $\{~ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} ~\}$.\\
\hstep{} is made of two sub-steps: a call to the meta language
unification and a check for \progress{} on the set of links, that intuitively
will compensate for the weaker equational theory honoured by \Ue.
\hrun{} compiles all terms in \foUnifPb{}, then executes each step and
finally decompiles the solution.
We claim:

@@

\noindent
That is, the two executions give the same result if all terms in
\foUnifPb are in the pattern fragment. Moreover:

@@

In other words if the two terms involved in a step
re-enter \llambda, then \hstep and \fstep are again related, even if
$\foUnifPb \not\subseteq \llambda$ and hence \cref{prop:fidelity} does not apply.
Indeed, the main difference between \cref{prop:fidelity} and \cref{prop:fidelity-recovery}
is that the assumption of the former is purely static, it can be checked upfront. 
When this assumption is not satisfied one can still simulate a logic program and
have guarantees of fidelity if, at run time, decidability of higher-order
unification is restored.

This property has a practical relevance since in many logic programming
implementations, including Elpi, the order in which unification problems
are tackled does matter.
The simplest example is the sequence $F \UnifRel \lambda x.a$ and
$F \appsep a \UnifRel a$: the second problem is not in \llambda and has two
unifiers, namely $\sigma_1 = \{~ F \mapsto \lambda x.x ~\}$ and
$\sigma_2 = \{~ F \mapsto \lambda x.a ~\}$. The first problem picks $\sigma_2$
making the second problem re-enter \llambda.

\paragraph{Backtracking} We omit it from our model of a logic programs execution
since it pays a very minor role, orthogonal to higher-order unification.
We point out that each \emph{run} corresponds to a (proof search) branch in the
logic program that either fails at some point, or succeeds. A computation that
succeeds by backtracking, exploring multiple branches, could be
modeled as set of runs with (possibly non empty) common prefixes.

\subsection{The solution (in a nutshell)}
\label{sec:nutshell}
A term $s$ is compiled to a term $t$ where every
``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
with an accessory \emph{link} that represents a suspended unification problem
$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, in the sense that
it does not contradict \Eo as it would otherwise do on the
``problematic'' sub-terms.\\
We now define ``problematic'' and ``well behaved'' more formally
We use the $\Diamond$ symbol since it stands for ``possibly'' in modal logic
and all problematic terms are characterized by some ``uncertainty''.

@@

\noindent
In other words the compiler outputs terms in \wellb, even if its
input is not.
Note that the property holds for any substitution. $\rho$ could be given by an
oracle and/or not necessarily be a most general one: in \wellb \Ue{} simply
does not contradict \Eo.
